---
title: "SOE forage fish index re-run"
format:
  html:
    code-fold: true
---

## Code

The input data was: `forageindex/fhdat/bluepyagg_stn_all_OISST_1982-2023.rds`, which was created at: `forageindex/fhdat/VASTforage_ProcessInputDat.R`
```{r, code = readLines(here::here("fhdat/VASTforage_ProcessInputDat.R")), eval = FALSE}
```


The forage fish VAST model was run with the code at: `forageindex/test_SOEpyindex/VASTunivariate_seasonalforageindex_operational.R`

```{r, code = readLines(here::here("test_SOEpyindex/VASTunivariate_seasonalforageindex_operational.R")), eval = FALSE}
```

The indicators were computed with the code at: `forageindex/test_SOEpyindex/SOE-VASTForageIndices.R`
```{r, code = readLines(here::here("test_SOEpyindex/SOE-VASTForageIndices.R")), eval = FALSE}
```

## Data comparison

New data:
```{r}
#| code-fold: false
spring <- readRDS(here::here("test_SOEpyindex/springforageindex.rds"))
fall <- readRDS(here::here("test_SOEpyindex/fallforageindex.rds"))
```

Original data:
```{r}
#| code-fold: false
orig_fall <- readRDS(here::here("SOEpyindex/fallforageindex.rds"))
orig_spring <- readRDS(here::here("SOEpyindex/springforageindex.rds"))
```

Combine and compare:
```{r}
#| code-fold: false
all_dat <- dplyr::bind_rows(spring |> dplyr::mutate(Season = "Spring", Origin = "New"),
                            fall |> dplyr::mutate(Season = "Fall", Origin = "New"),
                            orig_spring |> dplyr::mutate(Season = "Spring", Origin = "Original"),
                            orig_fall |> dplyr::mutate(Season = "Fall", Origin = "Original"))

all_dat |>
  ggplot2::ggplot(ggplot2::aes(x = Time,
                               y = Value,
                               color = Origin)) +
  ggplot2::geom_point() +
  ggplot2::geom_line() +
  ggplot2::facet_grid(cols = ggplot2::vars(EPU),
                      rows = ggplot2::vars(Var |> 
                                             stringr::str_remove_all(" Forage Fish Biomass Estimate")),
                      scales = "free_y") +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 30,
                                                     hjust = 1))
```
