---
title: "Assessing small pelagic fish trends in space and time using piscivore stomach contents: Supplement 4, Sensitivities"
author: "Sarah Gaichas, James Gartland, Brian Smith, Elizabeth Ng, Michael Celestino,
  Anthony Wood, Katie Drew, Abigail Tyrell, and James Thorson"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2: 
    toc: true
    toc_float: true
  bookdown::word_document2:
    toc: false
  bookdown::pdf_document2:
    includes: 
       in_header: latex/header1.tex
    keep_tex: true
link-citations: yes
csl: "canadian-journal-of-fisheries-and-aquatic-sciences.csl"
bibliography: FishDiet_EcoIndicators.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, # no code blocks in word doc
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
theme_set(theme_bw())
library(here)

```

# Forage index sensitivity to included prey species {-}

## Different prey cut-off {-}

Two methods were applied to derive the bluefish prey list. Initially, a cut-off of 10 observations of a prey item across all sampled bluefish stomachs was applied separately to the NEFSC food habits data and the NEAMAP food habits data. This approach resulted in the same common bluefish prey items for each dataset, but slightly different uncommon prey items. For example, Atlantic herring, Atlantic mackerel, and silver hake were included as prey items for NEFSC data, but not for NEAMAP data using this approach, due to the timing of the surveys combined with the distribution of large and small bluefish and these prey items inshore versus offshore. 

The second method (the one used for this paper) first combined the prey observations across NEFSC and NEAMAP food habits datasets and then applied a cutoff of 20 observations of a prey item to the combined dataset. This ensured that a consistent set of prey were used across predators for both datasets. The advantage of this approach is that any Atlantic herring, Atlantic mackerel, and silver hake seen in non-bluefish predator stomachs in the NEAMAP dataset would still be included in the prey index to reflect these prey nearshore. Further, categories reflecting different survey naming conventions for the same prey were combined using this approach (e.g., *Illex illecebrosus* and *Illex* sp.)

While the second method is appealing in its consistent treatment of prey items across the two datasets, the resulting index showed extremely low sensitivity to this change (dashed line compared with solid line below). This suggests that the index is primarily driven by the most commonly identified prey items across the datasets.  

```{r}
# compare 2022 prey list to paper update preylist for indices

stratlook <- data.frame(Stratum = c("Stratum_1",
                                    "Stratum_2",
                                    "Stratum_3",
                                    "Stratum_4",
                                    "Stratum_5",
                                    "Stratum_6",
                                    "Stratum_7",
                                    "Stratum_8",
                                    "Stratum_9",
                                    "Stratum_10",
                                    "Stratum_11",
                                    "Stratum_12",
                                    "Stratum_13",
                                    "Stratum_14",
                                    "Stratum_15"),
                        Region  = c("AllEPU", 
                                    "MABGB", 
                                    "MABGBstate", 
                                    "MABGBfed", 
                                    "MAB",
                                    "GB",
                                    "GOM",
                                    "bfall",
                                    "bfin",
                                    "bfoff",
                                    "MABGBalbinshore",
                                    "MABGBothoffshore",
                                    "albbfin",
                                    "albbfall",
                                    "allother"))

splitoutold <- read.csv("pyindex/2022/allagg_fall_500_lennosst_ALLsplit_biascorrect/Index.csv")
splitoutnew <- read.csv("pyindex/allagg_fall_500_lennosst_ALLsplit_biascorrect/Index.csv")

splitoutput2022 <- splitoutold %>%
  dplyr::left_join(stratlook) %>%
  dplyr::filter(Region %in% c("GOM", "GB", "MAB","MABGBstate", "bfin")) %>%
  dplyr::mutate(Type = ifelse(Region %in% c("GOM", "GB", "MAB"), "Ecoregion", "Bluefish"),
                Region = case_when(Region == "MABGBstate" ~ "StateWaters",
                                   Region == "bfin" ~ "SurveyBluefish",
                                   TRUE ~ Region))

splitoutput <- splitoutnew %>%
  dplyr::left_join(stratlook) %>%
  dplyr::filter(Region %in% c("GOM", "GB", "MAB","MABGBstate", "bfin")) %>%
  dplyr::mutate(Type = ifelse(Region %in% c("GOM", "GB", "MAB"), "Ecoregion", "Bluefish"),
                Region = case_when(Region == "MABGBstate" ~ "StateWaters",
                                   Region == "bfin" ~ "SurveyBluefish",
                                   TRUE ~ Region))

foragemax <- max(splitoutput$Estimate) #scale to new prey list

fig <- ggplot(splitoutput, aes(x=Time, y=Estimate, colour=Region)) +
  geom_errorbar(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate))+
  geom_point()+
  geom_line()+
  facet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to maximum") 



EPU <- ggplot(splitoutput %>% dplyr::filter(Region %in% c("GOM", "GB", "MAB")), 
              aes(x=Time, y=Estimate, colour=Region)) +
  geom_errorbar(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate))+
  geom_point()+
  geom_line()+
  geom_line(data=splitoutput2022 %>% dplyr::filter(Region %in% c("GOM", "GB", "MAB")),
            aes(x=Time, y=Estimate, colour=Region), linetype = "dashed") +
  facet_wrap(~fct_relevel(Region, "GOM", "GB", "MAB"), scales = "free_y", ncol = 1) +
  scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to maximum") 

BF <- ggplot(splitoutput %>% dplyr::filter(Region %in% c("StateWaters", "SurveyBluefish")), 
             aes(x=Time, y=Estimate, colour=Region)) +
  geom_errorbar(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate))+
  geom_point()+
  geom_line()+
  geom_line(data=splitoutput2022 %>% dplyr::filter(Region %in% c("StateWaters", "SurveyBluefish")),
            aes(x=Time, y=Estimate, colour=Region), linetype = "dashed") +
  facet_wrap(~fct_relevel(Region, "StateWaters", "SurveyBluefish"), scales = "free_y", ncol = 1) +
  scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to maximum")

EPU

BF

```


## Exclude assessed prey (Atlantic herring, Atlantic mackerel, silver hake) {-}

For a different application, three assessed prey were excluded from the dataset: Atlantic herring, Atlantic mackerel, and silver hake. In NEFSC stomach data, unidentified Clupeids are also likely to be Atlantic herring, so this prey group was excluded as well. Using this dataset, we can evaluate the sensitivity of the index to the prey list. We hypothesized that excluding these three assessed prey might make the most difference at the larger spatial scales including offshore regions (EPUs, Gulf of Maine, Georges Bank, and Mid-Atlantic). In particular, the dominance of silver hake in stomach content inputs from the Gulf of Maine and of silver hake and herring in stomach content inputs on Georges Bank noted in the manuscript suggests that excluding these species in those regions should affect the index.

The NEAMAP dataset had few of these three species observed in predator stomach contents sampled nearshore in the Mid-Atlantic bight, therefore we hypothesized that the index might be less sensitive to inclusion or exclusion at the scale of the Bluefish assessment indices.

Below we add the indices without Atlantic herring, Atlantic mackerel, and silver hake in dotted lines for comparison with the index using the initial prey list (dashed) and the full list included in this paper (solid line with error bars).

As hypothesized, there are large differences in both scale and trend for prey indices estimated by VAST for the Gulf of Maine and Georges Bank when excluding assessed prey species. There is a difference in scale for the Mid Atlantic, but index trends correspond better in that region.

Estimated indices at the bluefish assessment scale (state waters and bluefish inshore survey strata) have remarkably similar trends, although leaving out the three assessed prey species reduces the scale of the index.


```{r}
splitoutunmod <- read.csv("pyindex/unmod/allagg_fall_500_lennosst_ALLsplit_biascorrect/Index.csv")

splitoutputunmod <- splitoutunmod %>%
  dplyr::left_join(stratlook) %>%
  dplyr::filter(Region %in% c("GOM", "GB", "MAB","MABGBstate", "bfin")) %>%
  dplyr::mutate(Type = ifelse(Region %in% c("GOM", "GB", "MAB"), "Ecoregion", "Bluefish"),
                Region = case_when(Region == "MABGBstate" ~ "StateWaters",
                                   Region == "bfin" ~ "SurveyBluefish",
                                   TRUE ~ Region))

EPU <- ggplot(splitoutput %>% dplyr::filter(Region %in% c("GOM", "GB", "MAB")), 
              aes(x=Time, y=Estimate, colour=Region)) +
  geom_errorbar(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate))+
  geom_point()+
  geom_line()+
  geom_line(data=splitoutput2022 %>% dplyr::filter(Region %in% c("GOM", "GB", "MAB")),
            aes(x=Time, y=Estimate, colour=Region), linetype = "dashed") +
  geom_line(data=splitoutputunmod %>% dplyr::filter(Region %in% c("GOM", "GB", "MAB")),
            aes(x=Time, y=Estimate, colour=Region), linetype = "dotted") +
  facet_wrap(~fct_relevel(Region, "GOM", "GB", "MAB"), scales = "free_y", ncol = 1) +
  scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to maximum") 

BF <- ggplot(splitoutput %>% dplyr::filter(Region %in% c("StateWaters", "SurveyBluefish")), 
             aes(x=Time, y=Estimate, colour=Region)) +
  geom_errorbar(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate))+
  geom_point()+
  geom_line()+
  geom_line(data=splitoutput2022 %>% dplyr::filter(Region %in% c("StateWaters", "SurveyBluefish")),
            aes(x=Time, y=Estimate, colour=Region), linetype = "dashed") +
  geom_line(data=splitoutputunmod %>% dplyr::filter(Region %in% c("StateWaters", "SurveyBluefish")),
            aes(x=Time, y=Estimate, colour=Region), linetype = "dotted") +
  facet_wrap(~fct_relevel(Region, "StateWaters", "SurveyBluefish"), scales = "free_y", ncol = 1) +
  scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to maximum")

EPU

BF

```


# Forage index sensitivity to included predator species {-}

# VAST model fit and index sensitivity with depth as a covariate {-}

Comparisons are presented for models with and without depth as a covariate. This sensitivity was run for the fall model only, and include the "base" model (no covariates) and the two best fit models across all seasons (mean predator length, number of predators, and sea surface temperature as covariates, "lennumsst"; and mean predator length and number of predators as covariates "lennum") compared with depth only as a covariate, and depth combined with the two best fit models across all seasons: "deplennum" and "deplennumsst". 

```{r}

# from each output folder in pyindex, 
outdir <- here("pyindex_modsel2")
moddirs <- list.dirs(outdir) 
moddirs <- moddirs[-1]
# keep folder name
modnames <- list.dirs(outdir, full.names = FALSE)


# function to apply extracting info
getmodinfo <- function(d.name){
  # read settings
  modpath <- stringr::str_split(d.name, "/", simplify = TRUE)
  modname <- modpath[length(modpath)]
  
  settings <- read.table(file.path(d.name, "settings.txt"), comment.char = "",
    fill = TRUE, header = FALSE)
  
  n_x <- as.numeric(as.character(settings[(which(settings[,1]=="$n_x")+1),2]))
  grid_size_km <- as.numeric(as.character(settings[(which(settings[,1]=="$grid_size_km")+1),2]))
  max_cells <- as.numeric(as.character(settings[(which(settings[,1]=="$max_cells")+1),2]))
  use_anisotropy <- as.character(settings[(which(settings[,1]=="$use_anisotropy")+1),2])
  fine_scale <- as.character(settings[(which(settings[,1]=="$fine_scale")+1),2])
  bias.correct <- as.character(settings[(which(settings[,1]=="$bias.correct")+1),2])
  
  #FieldConfig
  if(settings[(which(settings[,1]=="$FieldConfig")+1),1]=="Component_1"){
    omega1 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+2),2])
    omega2 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+3),1])
    epsilon1 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+4),2])
    epsilon2 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+5),1])
    beta1 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+6),2])
    beta2 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+7),1])
  }
  
  if(settings[(which(settings[,1]=="$FieldConfig")+1),1]=="Omega1"){
    omega1 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+3),1])
    omega2 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+4),1])
    epsilon1 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+3),2])
    epsilon2 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+4),2])
    beta1 <- "IID"
    beta2 <- "IID"
  }
  
  
  #RhoConfig
  rho_beta1 <- as.numeric(as.character(settings[(which(settings[,1]=="$RhoConfig")+3),1]))
  rho_beta2 <- as.numeric(as.character(settings[(which(settings[,1]=="$RhoConfig")+3),2]))
  rho_epsilon1 <- as.numeric(as.character(settings[(which(settings[,1]=="$RhoConfig")+4),1]))
  rho_epsilon2 <- as.numeric(as.character(settings[(which(settings[,1]=="$RhoConfig")+4),2]))
  
  # read parameter estimates, object is called parameter_Estimates
  load(file.path(d.name, "parameter_estimates.RData"))
  
  AIC <- parameter_estimates$AIC[1]  
  converged <- parameter_estimates$Convergence_check[1]
  fixedcoeff <- unname(parameter_estimates$number_of_coefficients[2])
  randomcoeff <- unname(parameter_estimates$number_of_coefficients[3])
  
  #index <- read.csv(file.path(d.name, "Index.csv"))
  
  
  # return model attributes as a dataframe
  out <- data.frame(modname = modname,
                    n_x = n_x,
                    grid_size_km = grid_size_km,
                    max_cells = max_cells,
                    use_anisotropy = use_anisotropy,
                    fine_scale =  fine_scale,
                    bias.correct = bias.correct,
                    omega1 = omega1,
                    omega2 = omega2,
                    epsilon1 = epsilon1,
                    epsilon2 = epsilon2,
                    beta1 = beta1,
                    beta2 = beta2,
                    rho_epsilon1 = rho_epsilon1,
                    rho_epsilon2 = rho_epsilon2,
                    rho_beta1 = rho_beta1,
                    rho_beta2 = rho_beta2,
                    AIC = AIC,
                    converged = converged,
                    fixedcoeff = fixedcoeff,
                    randomcoeff = randomcoeff#,
                    #index = index
  )
  	
	return(out)

}


modcompare <- purrr::map_dfr(moddirs, getmodinfo)
```


For model comparisons with different combinations of catchability covariates, "modname" in \@ref(tab:modsel2) follows a similar pattern as in Supplement 2, with  all prey aggregated ("allagg" for all models), season ("fall" for models of months 7-12), number of knots (500 for all models), and which covariates were included. The names for model options and associated VAST model settings are: 

Model covariates sensitivity options, FieldConfig default (all IID), with anisotropy:

1.  "_base" =          No vessel overdispersion or length/number covariates 
<!--
1.  "_len" =           Predator mean length covariate
1.  "_num" =           Number of predator species covariate
-->
1.  "_lennum" =        Predator mean length and number of predator species covariates
<!--
1.  "_sst" =           Combined in situ and OISST covariate
1.  "_lensst" =        Predator mean length and SST covariates
1.  "_numsst" =        Number of predator species and SST covariates
-->
1.  "_lennumsst" =     Predator mean length, number of predator species, and SST covariates
<!--
1.  "_eta10" =         Overdispersion (vessel effect) in first linear predictor (prey encounter)
1.  "_eta11" =         Overdispersion (vessel effect) in both linear predictors (prey encounter and weight)
-->
1.  "_depth" =         Survey station depth covariate
1.  "_deplennum" =     Survey station depth, predator mean length and number of predator species covariates
1.  "_deplennumsst" =  Survey station depth, predator mean length, number of predator species, and SST covariates


```{r modsel2}
# from each output folder in pyindex, 
outdir <- here("pyindex_modsel2")
moddirs <- list.dirs(outdir) 
moddirs <- moddirs[-1]
# keep folder name
modnames <- list.dirs(outdir, full.names = FALSE)

modcompare2 <- purrr::map_dfr(moddirs, getmodinfo)

modselect2 <- modcompare2 %>%
  dplyr::mutate(season = case_when(str_detect(modname, "_fall_") ~ "Fall",
                            str_detect(modname, "spring") ~ "Spring",
                            str_detect(modname, "_all_") ~ "Annual",
                            TRUE ~ as.character(NA))) %>%
  dplyr::mutate(converged2 = case_when(str_detect(converged, "no evidence") ~ "likely",
                                str_detect(converged, "is likely not") ~ "unlikely",
                                TRUE ~ as.character(NA))) %>%
  dplyr::group_by(season) %>%
  dplyr::mutate(deltaAIC = AIC-min(AIC)) %>%
  dplyr::select(modname, season, deltaAIC, fixedcoeff,
         randomcoeff, use_anisotropy, 
         omega1, omega2, epsilon1, epsilon2, 
         beta1, beta2, AIC, converged2) %>%
  dplyr::arrange(season, AIC)

# DT::datatable(modselect2, rownames = FALSE, 
#               options= list(pageLength = 50, scrollX = TRUE),
#               caption = "Comparison of delta AIC values for alternative vessel effects and catchability covariates. See text for model descriptions.")

flextable::flextable(modselect2%>%
                       dplyr::select(-c(use_anisotropy, 
         omega1, omega2, epsilon1, epsilon2, 
         beta1, beta2))) %>%
  flextable::set_caption("Comparison of delta AIC values for alternative vessel effects and catchability covariates. See text for model descriptions.") %>%
  flextable::fontsize(size = 9, part = "all") %>%
  flextable::autofit()


```


Model comparisons above for fall reinforce initial model selection findings of the best model fit including mean predator length, number of predator species, and SST at a survey station as catchability covariates. Including depth in addition to these covariates resulted in slightly improved model fits. However, models with depth included as a covariate had convergence issues, with maximum gradients 

Indices resulting from the best fit models including/not including depth differ in scale, but differences in trend are less apparent.


```{r}
stratlook <- data.frame(Stratum = c("Stratum_1",
                                    "Stratum_2",
                                    "Stratum_3",
                                    "Stratum_4",
                                    "Stratum_5",
                                    "Stratum_6",
                                    "Stratum_7",
                                    "Stratum_8",
                                    "Stratum_9",
                                    "Stratum_10",
                                    "Stratum_11",
                                    "Stratum_12",
                                    "Stratum_13",
                                    "Stratum_14",
                                    "Stratum_15"),
                        Region  = c("AllEPU", 
                                    "MABGB", 
                                    "MABGBstate", 
                                    "MABGBfed", 
                                    "MAB",
                                    "GB",
                                    "GOM",
                                    "bfall",
                                    "bfin",
                                    "bfoff",
                                    "MABGBalbinshore",
                                    "MABGBothoffshore",
                                    "albbfin",
                                    "albbfall",
                                    "allother"))


```


```{r indexcompare}

# function to apply extracting info
getmodindex <- function(d.name){
  # read settings
  modpath <- stringr::str_split(d.name, "/", simplify = TRUE)
  modname <- modpath[length(modpath)]
  
  index <- read.csv(file.path(d.name, "Index.csv"))
  # return model indices as a dataframe
  out <- data.frame(modname = modname,
                    index
  )
  
  return(out)
}

modcompareindex <- purrr::map_dfr(moddirs, getmodindex)

splitout <- modcompareindex #read.csv("pyindex/allagg_spring_500_lennosst_ALLsplit_biascorrect/Index.csv")

splitoutput <- splitout %>%
  dplyr::left_join(stratlook) %>%
  dplyr::filter(Region %in% c("GOM", "GB", "MAB","MABGBstate", "bfin")) %>%
  dplyr::mutate(Type = ifelse(Region %in% c("GOM", "GB", "MAB"), "Ecoregion", "Bluefish"),
                Region = case_when(Region == "MABGBstate" ~ "StateWaters",
                                   Region == "bfin" ~ "SurveyBluefish",
                                   TRUE ~ Region))

foragemax <- max(splitoutput$Estimate)

ggplot(splitoutput, aes(x=Time, y=Estimate, colour=modname)) +
  geom_errorbar(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate))+
  geom_point()+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y") +
  scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to maximum")  
  



```
```{r}

foragetsmean <- splitoutput %>%
  dplyr::group_by(modname, Region) %>%
  dplyr::mutate(fmean = mean(Estimate)) 


ggplot(foragetsmean, aes(x=Time, y=((Estimate-fmean)/fmean), colour=modname)) +
  #geom_errorbar(aes(ymin=(Estimate+Std..Error.for.Estimate - fmean)/fmean, 
  #                  ymax=(Estimate-Std..Error.for.Estimate - fmean)/fmean))+
  geom_point()+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y") +
  #scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to mean")  

```
Additional work to evaluate why the models with a depth covariate were less likely to converge is warranted if information on scale is important to the index. For our application as a catchability covariate for a stock assessment index, scale was less important than trend. Therefore, the fully converged models without depth as a covariate were put forward as the final index.

# References {-}