---
title: "Assessing small pelagic fish trends in space and time using piscivore stomach contents: Supplement 1, SST Covariate Data and Index Sensitivity"
author: "Sarah Gaichas, James Gartland, Brian Smith, Anthony Wood, Elizabeth Ng, Michael Celestino,
  Katie Drew, Abigail Tyrell, and James Thorson"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::word_document2:
    toc: false
  bookdown::html_document2: 
    toc: true
    toc_float: true
  bookdown::pdf_document2:
    includes: 
       in_header: latex/header1.tex
    keep_tex: true
link-citations: yes
csl: "canadian-journal-of-fisheries-and-aquatic-sciences.csl"
bibliography: FishDiet_EcoIndicators.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, # no code blocks in word doc
                      message = FALSE,
                      warning = FALSE,
                      fig.align = 'center')

library(tidyverse)
theme_set(theme_bw())
library(here)
library(FishStatsUtils)

```

The number of survey stations missing surface temperature data varied considerably by decade. A large percentage of survey stations lacked in-situ temperature measurements between 1985 and 1990, while the percentage of stations missing temperature was generally below 10% (with a few exceptions) from 1991-2021 (Table \@ref(tab:sstdat)). Therefore, OISST temperature estimates were more commonly substituted early in the time series. 

```{r sstdat}
#[add stuff about how many SST values were missing and which years got substituted]

inputdat <- readRDS(here("fhdat/bluepyagg_stn_all_OISST.rds"))

stnsummary <- inputdat %>%
  dplyr::filter(year>1984) %>%
  dplyr::group_by(year, season_ng) %>%
  dplyr::summarise(nstation = n(),
                   hastemp = sum(!is.na(surftemp))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(percentmiss = (nstation - hastemp)/nstation*100) %>%
  tidyr::pivot_wider(names_from = season_ng,
    values_from = c(nstation, hastemp, percentmiss)) %>%
  dplyr::select(year, nstation_SPRING, hastemp_SPRING, percentmiss_SPRING,
                nstation_FALL, hastemp_FALL, percentmiss_FALL)

flextable::flextable(stnsummary) %>%
  flextable::set_header_labels(year = 'Year', 
                               nstation_SPRING = "Spring N stations", 
                               hastemp_SPRING = "Spring N stations with in situ temperature",
                               percentmiss_SPRING = "Spring percent missing temperature",
                               nstation_FALL = "Fall N stations", 
                               hastemp_FALL = "Fall N stations with in situ temperature",
                               percentmiss_FALL = "Fall percent missing temperature") %>%
  flextable::set_caption("Number of survey stations by year and season with in situ sea surface temperature measurements.") %>%
  flextable::colformat_double(big.mark = "", digits = 0) %>%
  #flextable::autofit()
  flextable::width(width = 1)

```

## Comparison of survey in situ surface temperature and OISST {-}

For stations with in-situ surface temperature measurements, we compared these to OISST from the area and date of each survey station. The spatial resolution of the OISST dataset is 25 km, so survey stations anywhere within the 25 km block would be assigned the same temperature. 

Spatial resolution is visible in a sample plot of OISST from March 15, 2021 (Fig. \@ref(fig:sampleOISST)).

```{r sampleOISST,  fig.cap="March 15 2021 OISST for the Northeast US Shelf."}
#fig.show='hold', out.width='49%',
#visualize
sst2021 <- readRDS(here("data-raw/gridded/sst_data/sst2021.rds"))

oneday <- sst2021[sst2021$month=="07" & sst2021$day=="04",] 

jan15 <- sst2021[sst2021$month=="01" & sst2021$day=="15",] 

mar15 <- sst2021[sst2021$month=="03" & sst2021$day=="15",]

may15 <- sst2021[sst2021$month=="05" & sst2021$day=="15",]

jul15 <- sst2021[sst2021$month=="07" & sst2021$day=="15",]

sep15 <- sst2021[sst2021$month=="09" & sst2021$day=="15",]

nov15 <- sst2021[sst2021$month=="11" & sst2021$day=="15",]

dailysstplot <- function(oneday){
  ggplot() +
    geom_tile(data = oneday, aes(x = Lon, y = Lat, fill = sst)) +
    geom_sf(data = ecodata::coast) +
    geom_point(data = northwest_atlantic_grid, aes(x = Lon, y = Lat), size=0.05, alpha=0.1) +
    scale_fill_gradientn(name = "Temp C",
                         limits = c(0.5, 31),
                         colours = c(scales::muted("blue"), "white",
                                     scales::muted("red"), "black")) +
    coord_sf(xlim = c(-77, -65), ylim = c(35, 45)) + 
    ggtitle(paste("SST, mm dd yyyy:", unique(oneday$month),
                   unique(oneday$day), unique(oneday$year), sep = " "))
}

#dailysstplot(oneday)

#par(mfrow=c(2,3))
#dailysstplot(jan15)
dailysstplot(mar15)
#dailysstplot(may15)
#dailysstplot(jul15)
#dailysstplot(sep15)
#dailysstplot(nov15)
```

The range of temperatures observed on surveys differs by season (Fig. \@ref(fig:temprange)). 

```{r temprange, fig.cap = "Range of in-situ surface temperatures from the NEFSC bottom trawl surveys by season, 1985-2021."}

insitusst <- inputdat |>
  dplyr::filter(year>1984) |>
  dplyr::select(surftemp, season_ng) |>
  na.omit()
  
ggplot2::ggplot(insitusst, aes(x=surftemp)) + 
  geom_histogram() + 
  facet_wrap(~season_ng, ncol = 1)

```


In situ survey surface temperatures compare well with OISST where there are both available over a wide range of temperatures, with some exceptions (Fig. \@ref(fig:comparestations)). 

```{r comparestations, fig.cap="Comparison of NEFSC in situ surface temperature measurements (surftemp) with OISST temperatures (oisst) for the same date and 25 km grid cell. The blue line is a smooth through the data, the black line is the desired 1:1 relationship where temperatures match."}


comparesst <- inputdat %>%
  dplyr::filter(year>1984)%>%
  dplyr::select(surftemp, oisst) %>%
  na.omit()

ggplot2::ggplot(comparesst, aes(x=surftemp, y=oisst)) +
  geom_point(color="darkgray", alpha=0.3)+
  geom_abline(intercept = 0, slope = 1) +
  geom_smooth() +
  theme_bw() 
```

```{r}
mapsst <- inputdat %>%
  dplyr::filter(year>1984) %>%
  dplyr::mutate(sstdiff = surftemp-oisst) %>%
  dplyr::select(id, year, season_ng, declon, declat, surftemp, oisst, sstdiff) 

yrmap <- function(mapyr){
  ggplot2::ggplot(mapsst%>%filter(year==mapyr)) +
  geom_sf(data = ecodata::coast) +
  coord_sf(xlim = c(-77, -65), ylim = c(35, 45)) + 
  geom_point(aes(x=declon, y=declat, colour=sstdiff)) +
  scale_color_gradient2(low = "blue",
                        mid = "green",
                        high = "purple",
                        midpoint = 0,
                        na.value = "yellow") +
  theme_bw() +
  facet_wrap(~season_ng) +
  ggtitle(paste("SST difference survey-OISST:", mapyr, sep = " "))
}
```


In the following maps of survey station locations by SST match/mismatch for each year, yellow points indicate missing data that would be filled by OISST, and the color range shows how different in situ and OISST temperature is for stations with both values (Fig. \@ref(fig:matchmaps)).

Most mismatches were near Cape Hatteras where the temperature gradient from the nearby Gulf Stream is not well captured by the resolution of OISST. Others were similarly near the continental slope.

```{r matchmaps,  results='asis', fig.cap="Comparison of in-situ survey surface temperature and OISST for selected years."}
#fig.show='hold',

i=1

for(mapyr in c(1985, 1990, 1995, 2000, 2005, 2010, 2015, 2020)){
  
    cat("  \n",  letters[i],"  \n")
    print(yrmap(mapyr)) 
    cat("  \n") 
    i <- i+1
    
  }
```


Despite some mismatches, we felt substituting OISST temperature to retain the stomach samples at stations with missing temperature data was preferable to leaving stomach data out due to missing covariate information.

## Index sensitivity to SST covariates  {-}

The two best fitting models both include mean predator length and number of predator species as covariates; adding sea surface temperature to these covariates resulted in the best fitting model in both seasons. To evaluate the overall impact of SST as a covariate, we first compare the indices with and without this covariate (best and second best fit to the data, respectively, although in fall the difference in AIC was small). These indices are not bias corrected.

Index trends estimated by both models are nearly identical (\@ref(fig:trendcomp)). Including SST slightly increased index scale relative to the second best fitting model in fall, but had the opposite effect in spring (\@ref(fig:scalecomp)). Therefore, including OISST to fill in missing in-situ temperature data seems to have no impact on index trend (as the sst covariate itself has no impact on index trend). 

```{r}
stratlook <- data.frame(Stratum = c("Stratum_1",
                                    "Stratum_2",
                                    "Stratum_3",
                                    "Stratum_4",
                                    "Stratum_5",
                                    "Stratum_6",
                                    "Stratum_7",
                                    "Stratum_8",
                                    "Stratum_9",
                                    "Stratum_10",
                                    "Stratum_11",
                                    "Stratum_12",
                                    "Stratum_13",
                                    "Stratum_14",
                                    "Stratum_15"),
                        Region  = c("AllEPU", 
                                    "MABGB", 
                                    "MABGBstate", 
                                    "MABGBfed", 
                                    "MAB",
                                    "GB",
                                    "GOM",
                                    "bfall",
                                    "bfin",
                                    "bfoff",
                                    "MABGBalbinshore",
                                    "MABGBothoffshore",
                                    "albbfin",
                                    "albbfall",
                                    "allother"))


```

```{r trendcompfall, fig.cap="Trend comparisons between fall forage indices estimated with and without SST as a catchability covariate.",  fig.height=8}

# from each output folder in pyindex, 
outdir <- here("pyindex_modsel2")
moddirs <- list.dirs(outdir) 
moddirs <- moddirs[-1]
# keep folder name
#modnames <- list.dirs(outdir, full.names = FALSE)

#hack to get only lennum and lennumsst
moddirs <- moddirs[stringr::str_detect(moddirs, "lennum")]


# function to apply extracting info
getmodindex <- function(d.name){
  # read settings
  modpath <- stringr::str_split(d.name, "/", simplify = TRUE)
  modname <- modpath[length(modpath)]
  
  index <- read.csv(file.path(d.name, "Index.csv"))
  # return model indices as a dataframe
  out <- data.frame(modname = modname,
                    index
  )
  
  return(out)
}

modcompareindex <- purrr::map_dfr(moddirs, getmodindex)

splitoutput <- modcompareindex %>%
  dplyr::mutate(Season = modname |> map(str_split, pattern = "_") |> map_chr(c(1,2))) %>%
  dplyr::left_join(stratlook) %>%
  dplyr::filter(Region %in% c("GOM", "GB", "MAB","MABGBstate", "bfin")) %>%
  dplyr::mutate(Type = ifelse(Region %in% c("GOM", "GB", "MAB"), "Ecoregion", "Bluefish"),
                Region = case_when(Region == "MABGBstate" ~ "StateWaters",
                                   Region == "bfin" ~ "SurveyBluefish",
                                   TRUE ~ Region))

foragemax <- max(splitoutput$Estimate)

foragetsmean <- splitoutput %>%
  dplyr::group_by(modname, Region) %>%
  dplyr::mutate(fmean = mean(Estimate)) 


ggplot(foragetsmean |> filter(Season =="fall"), aes(x=Time, y=((Estimate-fmean)/fmean), colour=modname)) +
  #geom_errorbar(aes(ymin=(Estimate+Std..Error.for.Estimate - fmean)/fmean, 
  #                  ymax=(Estimate-Std..Error.for.Estimate - fmean)/fmean))+
  geom_point(aes(shape=modname))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  #scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to time series mean")  +
  ggtitle("Fall models: trend comparison") +
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0),
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))

```

```{r trendcompspring, fig.cap="Trend comparisons between spring forage indices estimated with and without SST as a catchability covariate.",  fig.height=8}


ggplot(foragetsmean |> filter(Season =="spring"), aes(x=Time, y=((Estimate-fmean)/fmean), colour=modname)) +
  #geom_errorbar(aes(ymin=(Estimate+Std..Error.for.Estimate - fmean)/fmean, 
  #                  ymax=(Estimate-Std..Error.for.Estimate - fmean)/fmean))+
  geom_point(aes(shape=modname))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  #scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to time series mean")  +
  ggtitle("Spring models: trend comparison") +
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0),
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))
```


```{r scalecompfall, fig.cap="Scale comparisons between fall forage indices estimated with and without SST as a catchability covariate.",  fig.height=8}
# from each output folder in pyindex, 

ggplot(splitoutput |> filter(Season =="fall"), aes(x=Time, y=Estimate, colour=modname)) +
  #geom_errorbar(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate))+
  geom_ribbon(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate, fill=modname), linetype = 0, alpha = 0.15)+
  geom_point(aes(shape=modname))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to maximum")  +
  ggtitle("Fall models: scale comparison")+
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0),
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))

```

```{r scalecompspring, fig.cap="Scale comparisons between spring forage indices estimated with and without SST as a catchability covariate.",  fig.height=8}

ggplot(splitoutput |> filter(Season =="spring"), aes(x=Time, y=Estimate, colour=modname)) +
  #geom_errorbar(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate))+
  geom_ribbon(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate, fill=modname), linetype = 0, alpha = 0.15)+
  geom_point(aes(shape=modname))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to maximum")  +
  ggtitle("Spring models: scale comparison")+
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0),
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))


```

