---
title: "Forage Index update for 2024 SOE"
subtitle: "Data updated through 2022"
author: "Sarah Gaichas, James Gartland, Brian Smith, Anthony Wood, Elizabeth Ng, Michael Celestino,
  Katie Drew, Abigail Tyrell, and James Thorson"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2: 
    toc: true
    toc_float: true
    code_fold: hide
  bookdown::word_document2:
    toc: true
  bookdown::pdf_document2:
    includes: 
       in_header: latex/header1.tex
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, # no code blocks in word doc
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
theme_set(theme_bw())
library(here)

library(officedown)
library(officer)
library(flextable)
# op_section <- prop_section(type = "continuous")
# close_section <- prop_section(
#     page_size = page_size(orient = "landscape"), 
#     type = "continuous")

```

# Summary

Based on comparisons below, I'm satisfied that the methods update between last years SOE and this doesn't fundamentally change the trend. I'm also fine with starting the time series earlier. However, looking at the coverage, we may not want to start the model at the absolute beginning of the time series.

Will discuss a good starting point with the team, but for now submit the 1982 starting point with data through 2022.

# Workflow

Operational updates to the forage index submitted to the State of the Ecosystem report are in the forageindex github repository: https://github.com/NOAA-EDAB/forageindex

Data input files are in the folder `fhdat` and were processed with the script `VASTforage_ProcessInputDat.R` in that folder: https://github.com/NOAA-EDAB/forageindex/blob/main/fhdat/VASTforage_ProcessInputDat.R

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/forageindex/main/fhdat/VASTforage_ProcessInputDat.R"), eval=FALSE}

```


VAST models were run using the script `VASTunivariate_seasonalforageindex_operational.R` in the folder `VASTscripts`: https://github.com/NOAA-EDAB/forageindex/blob/main/VASTscripts/VASTunivariate_seasonalforageindex_operational.R 

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/forageindex/main/VASTscripts/VASTunivariate_seasonalforageindex_operational.R"), eval=FALSE}

```

Model output was saved in the folder `SOEpyindex` for three different time series options: 1973-2022, 1982-2022, and 1985-2022.

The script to create the SOE forage indices from the VAST output is in the `SOEpyindex` folder: https://github.com/NOAA-EDAB/forageindex/blob/main/SOEpyindex/SOE-VASTForageIndices.R

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/forageindex/main/SOEpyindex/SOE-VASTForageIndices.R"), eval=FALSE}

```

Code blocks below setup the index comparisons in the remainder of this document.

```{r preycut1, fig.cap="Trend comparison between fall forage indices using different bluefish prey cut-offs: 10 or more observations by survey and 25 observations combined across surveys (current method).", fig.height=8}
# compare 2022 prey list to paper update preylist for indices

stratlook <- data.frame(Stratum = c("Stratum_1",
                                    "Stratum_2",
                                    "Stratum_3",
                                    "Stratum_4",
                                    "Stratum_5",
                                    "Stratum_6",
                                    "Stratum_7",
                                    "Stratum_8",
                                    "Stratum_9",
                                    "Stratum_10",
                                    "Stratum_11",
                                    "Stratum_12",
                                    "Stratum_13",
                                    "Stratum_14",
                                    "Stratum_15"),
                        Region  = c("AllEPU", 
                                    "MABGB", 
                                    "MABGBstate", 
                                    "MABGBfed", 
                                    "MAB",
                                    "GB",
                                    "GOM",
                                    "bfall",
                                    "bfin",
                                    "bfoff",
                                    "MABGBalbinshore",
                                    "MABGBothoffshore",
                                    "albbfin",
                                    "albbfall",
                                    "allother"))
```

```{r getmodinfoindex}

# from each output folder in pyindex, 
outdir <- here("SOEpyindex")
moddirs <- list.dirs(outdir)
# only get pred sensitivities and original full models
fullfall <- grep("/allagg_fall_", moddirs, value = TRUE)
fullspring <- grep("/allagg_spring_", moddirs, value = TRUE)
moddirs <- c(fullfall, fullspring)
# get original published version 1985-2021
outdir1 <- here("pyindex")
moddirs1 <- list.dirs(outdir1)
RTfall <- grep("pyindex/2022/allagg_fall_", moddirs1, value = TRUE)
RTspring <- grep("pyindex/2022/allagg_spring_", moddirs1, value = TRUE)
paperfall <- grep("pyindex/allagg_fall_", moddirs1, value = TRUE)
paperspring <- grep("pyindex/allagg_spring_", moddirs1, value = TRUE)
moddirs1 <- c(RTfall, RTspring, paperfall, paperspring)
# keep folder name
modnames <- list.dirs(outdir, full.names = FALSE)
modnames1 <- list.dirs(outdir1, full.names = FALSE)

# function to apply extracting info
getmodinfo <- function(d.name){
  # read settings
  modpath <- stringr::str_split(d.name, "/", simplify = TRUE)
  modname <- modpath[length(modpath)]
  modnamepre <- modpath[length(modpath)-1]
  
  settings <- read.table(file.path(d.name, "settings.txt"), comment.char = "",
    fill = TRUE, header = FALSE)
  
  n_x <- as.numeric(as.character(settings[(which(settings[,1]=="$n_x")+1),2]))
  grid_size_km <- as.numeric(as.character(settings[(which(settings[,1]=="$grid_size_km")+1),2]))
  max_cells <- as.numeric(as.character(settings[(which(settings[,1]=="$max_cells")+1),2]))
  use_anisotropy <- as.character(settings[(which(settings[,1]=="$use_anisotropy")+1),2])
  fine_scale <- as.character(settings[(which(settings[,1]=="$fine_scale")+1),2])
  bias.correct <- as.character(settings[(which(settings[,1]=="$bias.correct")+1),2])
  
  #FieldConfig
  if(settings[(which(settings[,1]=="$FieldConfig")+1),1]=="Component_1"){
    omega1 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+2),2])
    omega2 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+3),1])
    epsilon1 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+4),2])
    epsilon2 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+5),1])
    beta1 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+6),2])
    beta2 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+7),1])
  }
  
  if(settings[(which(settings[,1]=="$FieldConfig")+1),1]=="Omega1"){
    omega1 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+3),1])
    omega2 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+4),1])
    epsilon1 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+3),2])
    epsilon2 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+4),2])
    beta1 <- "IID"
    beta2 <- "IID"
  }
  
  
  #RhoConfig
  rho_beta1 <- as.numeric(as.character(settings[(which(settings[,1]=="$RhoConfig")+3),1]))
  rho_beta2 <- as.numeric(as.character(settings[(which(settings[,1]=="$RhoConfig")+3),2]))
  rho_epsilon1 <- as.numeric(as.character(settings[(which(settings[,1]=="$RhoConfig")+4),1]))
  rho_epsilon2 <- as.numeric(as.character(settings[(which(settings[,1]=="$RhoConfig")+4),2]))
  
  # read parameter estimates, object is called parameter_Estimates
  load(file.path(d.name, "parameter_estimates.RData"))
  
  AIC <- parameter_estimates$AIC[1]  
  converged <- parameter_estimates$Convergence_check[1]
  fixedcoeff <- unname(parameter_estimates$number_of_coefficients[2])
  randomcoeff <- unname(parameter_estimates$number_of_coefficients[3])
  
  #index <- read.csv(file.path(d.name, "Index.csv"))
  
  # return model attributes as a dataframe
  out <- data.frame(modname = paste0(modnamepre,"_", modname),
                    n_x = n_x,
                    grid_size_km = grid_size_km,
                    max_cells = max_cells,
                    use_anisotropy = use_anisotropy,
                    fine_scale =  fine_scale,
                    bias.correct = bias.correct,
                    omega1 = omega1,
                    omega2 = omega2,
                    epsilon1 = epsilon1,
                    epsilon2 = epsilon2,
                    beta1 = beta1,
                    beta2 = beta2,
                    rho_epsilon1 = rho_epsilon1,
                    rho_epsilon2 = rho_epsilon2,
                    rho_beta1 = rho_beta1,
                    rho_beta2 = rho_beta2,
                    AIC = AIC,
                    converged = converged,
                    fixedcoeff = fixedcoeff,
                    randomcoeff = randomcoeff#,
                    #index = index
  )
  	
	return(out)

}

# getfitdat <- function(d.name){
#   # read settings
#   modpath <- stringr::str_split(d.name, "/", simplify = TRUE)
#   modname <- modpath[length(modpath)]
#   
#   readRDS(file.path(d.name, "fit.rds"))
#   nstations <- fit$data_list$n_i # this is grid size not input N stations
#   out <- data.frame(modname = modname,
#                     nstations = nstations
#   )
#   
#   return(out)
# }


# function to apply extracting info
getmodindex <- function(d.name){
  # read settings
  modpath <- stringr::str_split(d.name, "/", simplify = TRUE)
  modname <- modpath[length(modpath)]
  modnamepre <- modpath[length(modpath)-1]
  
  index <- read.csv(file.path(d.name, "Index.csv"))
  # return model indices as a dataframe
  out <- data.frame(modname = paste0(modnamepre,"_", modname),
                    index
  )
  
  return(out)
}

moddirsall <- c(moddirs, moddirs1)

modcompare <- purrr::map_dfr(moddirsall, getmodinfo)

modcompareindex <- purrr::map_dfr(moddirsall, getmodindex)

```

# Trend comparisons

## Methods change and adding one year

Lets compare last year's SOE submission which was used in the 2022 Bluefish RT (2022) with the slight adjustment in prey list for the published CJFAS paper (pyindex) and with the same methods as in the CJFAS paper but extending data by 1 year to 2022.

This is fall trend.

```{r comptrend1, fig.cap="Trend comparison between fall forage indices using 1985-2021 data for 2022 Bluefish RT and 2023 SOE (2022), CJFAS published (pyindex), and CJFAS published index updated to end in 2022 (1985-2022).", fig.height=8}
splitoutput <- modcompareindex %>%
  dplyr::mutate(Season = modname |> map(str_split, pattern = "_") |> map_chr(c(1,3))) %>%
  dplyr::mutate(Data = modname |> map(str_split, pattern = "_") |> map_chr(c(1,1))) %>% 
  dplyr::left_join(stratlook) %>%
  dplyr::filter(Region %in% c("GOM", "GB", "MAB","MABGBstate", "bfin")) %>%
  dplyr::mutate(Type = ifelse(Region %in% c("GOM", "GB", "MAB"), "Ecoregion", "Bluefish"),
                Region = case_when(Region == "MABGBstate" ~ "StateWaters",
                                   Region == "bfin" ~ "SurveyBluefish",
                                   TRUE ~ Region))

foragemax <- max(splitoutput$Estimate)


foragetsmean <- splitoutput %>%
  dplyr::group_by(modname, Region) %>%
  dplyr::mutate(fmean = mean(Estimate)) 


ggplot(foragetsmean |> filter(Season =="fall",
                             Data %in% c("2022",
                                         "pyindex",
                                         "1985-2022")), 
       aes(x=Time, y=((Estimate-fmean)/fmean), colour=Data)) +
  #geom_errorbar(aes(ymin=(Estimate+Std..Error.for.Estimate - fmean)/fmean, 
  #                  ymax=(Estimate-Std..Error.for.Estimate - fmean)/fmean))+
  geom_point(aes(shape=Data))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  #scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to time series mean")  +
  ggtitle("Fall models: trend comparison") +
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0)
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))

```

This is spring trend.

```{r comptrend2, fig.cap="Trend comparison between spring forage indices using 1985-2021 data for 2022 Bluefish RT and 2023 SOE (2022), CJFAS published (pyindex), and CJFAS published index updated to end in 2022 (1985-2022).", fig.height=8}

ggplot(foragetsmean |> filter(Season =="spring",
                             Data %in% c("2022",
                                         "pyindex",
                                         "1985-2022")), 
       aes(x=Time, y=((Estimate-fmean)/fmean), colour=Data)) +
  #geom_errorbar(aes(ymin=(Estimate+Std..Error.for.Estimate - fmean)/fmean,
  #                  ymax=(Estimate-Std..Error.for.Estimate - fmean)/fmean))+
  geom_point(aes(shape=Data))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  #scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to time series mean")  +
  ggtitle("Spring models: trend comparison")+
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0)
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))


```

## How far back can we go?

Spatial coverage for stomachs in fall

```{r}
mapfall <- paste0(moddirs[1], "/Data_by_year.png")

knitr::include_graphics(mapfall) 
```

Spatial coverage for stomachs in spring

```{r}
mapspring <- paste0(moddirs[4], "/Data_by_year.png")

knitr::include_graphics(mapspring) 
```
Now lets compare trends when we add 2022 and also extend the data back from the CJFAS dataset, with starting points at the beginning of stomach collections (1973-2022), at the earliest full OISST dataset for SST substitution (1982-2022), and the time series start used last year and for bluefish (1985-2022). 

This is fall.

```{r comptrend3, fig.cap="Trend comparison between fall forage indices using CJFAS published method with datasets from 1973-2022, 1982-2022, and 1985-2022.", fig.height=8}

ggplot(foragetsmean |> filter(Season =="fall",
                             Data %in% c("1973-2022",
                                         "1982-2022",
                                         "1985-2022")), 
       aes(x=Time, y=((Estimate-fmean)/fmean), colour=Data)) +
  #geom_errorbar(aes(ymin=(Estimate+Std..Error.for.Estimate - fmean)/fmean,
  #                  ymax=(Estimate-Std..Error.for.Estimate - fmean)/fmean))+
  geom_point(aes(shape=Data))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  #scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to time series mean")  +
  ggtitle("Fall models: trend comparison")+
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0)
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))


```


This is spring. 


```{r comptrend4, fig.cap="Trend comparison between spring forage indices using CJFAS published method with datasets from 1973-2022, 1982-2022, and 1985-2022.", fig.height=8}

ggplot(foragetsmean |> filter(Season =="spring",
                             Data %in% c("1973-2022",
                                         "1982-2022",
                                         "1985-2022")), 
       aes(x=Time, y=((Estimate-fmean)/fmean), colour=Data)) +
  #geom_errorbar(aes(ymin=(Estimate+Std..Error.for.Estimate - fmean)/fmean,
  #                  ymax=(Estimate-Std..Error.for.Estimate - fmean)/fmean))+
  geom_point(aes(shape=Data))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  #scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to time series mean")  +
  ggtitle("Spring models: trend comparison")+
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0)
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))


```

# Scale comparisons

We don't use scale much, but in case someone does here they are, scaled relative to the max observed across all datasets.

## Methods change and adding one year

Same comparisons as above: fall

```{r compscale1, fig.cap="Scale comparison between fall forage indices using 1985-2021 data for 2022 Bluefish RT and 2023 SOE (2022), CJFAS published (pyindex), and CJFAS published index updated to end in 2022 (1985-2022).", fig.height=8}


ggplot(splitoutput |> filter(Season =="fall",
                             Data %in% c("2022",
                                         "pyindex",
                                         "1985-2022")), 
       aes(x=Time, y=Estimate, colour=Data)) +
  #geom_errorbar(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate))+
  geom_ribbon(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate, fill=Data), linetype = 0, alpha = 0.15)+
  geom_point(aes(shape=Data))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to maximum")  +
  ggtitle("Fall models: scale comparison")+
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0)
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))

```

Spring

```{r compscale2, fig.cap="Scale comparison between spring forage indices using 1985-2021 data for 2022 Bluefish RT and 2023 SOE (2022), CJFAS published (pyindex), and CJFAS published index updated to end in 2022 (1985-2022).", fig.height=8}
  

ggplot(splitoutput |> filter(Season =="spring",
                             Data %in% c("2022",
                                         "pyindex",
                                         "1985-2022")), 
       aes(x=Time, y=Estimate, colour=Data)) +
  #geom_errorbar(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate))+
  geom_ribbon(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate, fill=Data), linetype = 0, alpha = 0.15)+
  geom_point(aes(shape=Data))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to maximum")  +
  ggtitle("Spring models: scale comparison")+
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0)
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))


```


## How far back can we go?

Same comparisons as above: fall

```{r compscale3, fig.cap="Scale comparison between fall forage indices using 1985-2021 data for 2022 Bluefish RT and 2023 SOE (2022), CJFAS published (pyindex), and CJFAS published index updated to end in 2022 (1985-2022).", fig.height=8}


ggplot(splitoutput |> filter(Season =="fall",
                             Data %in% c("1973-2022",
                                         "1982-2022",
                                         "1985-2022")), 
       aes(x=Time, y=Estimate, colour=Data)) +
  #geom_errorbar(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate))+
  geom_ribbon(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate, fill=Data), linetype = 0, alpha = 0.15)+
  geom_point(aes(shape=Data))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to maximum")  +
  ggtitle("Fall models: scale comparison")+
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0)
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))

```

Spring

```{r compscale4, fig.cap="Scale comparison between spring forage indices using 1985-2021 data for 2022 Bluefish RT and 2023 SOE (2022), CJFAS published (pyindex), and CJFAS published index updated to end in 2022 (1985-2022).", fig.height=8}
  

ggplot(splitoutput |> filter(Season =="spring",
                             Data %in% c("1973-2022",
                                         "1982-2022",
                                         "1985-2022")), 
       aes(x=Time, y=Estimate, colour=Data)) +
  #geom_errorbar(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate))+
  geom_ribbon(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate, fill=Data), linetype = 0, alpha = 0.15)+
  geom_point(aes(shape=Data))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to maximum")  +
  ggtitle("Spring models: scale comparison")+
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0)
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))


```

