---
title: "Forage Index update for 2025 SOE"
subtitle: "Data updated through 2023"
author: "Sarah Gaichas, James Gartland, Brian Smith, Anthony Wood, Elizabeth Ng, Michael Celestino,
  Katie Drew, Abigail Tyrell, and James Thorson"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2: 
    toc: true
    toc_float: true
    code_fold: hide
  bookdown::word_document2:
    toc: true
  bookdown::pdf_document2:
    includes: 
       in_header: latex/header1.tex
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, # no code blocks in word doc
                      message = FALSE,
                      warning = FALSE)

library(tidyverse)
theme_set(theme_bw())
library(here)

library(officedown)
library(officer)
library(flextable)
# op_section <- prop_section(type = "continuous")
# close_section <- prop_section(
#     page_size = page_size(orient = "landscape"), 
#     type = "continuous")

```

# Summary

I needed to compare the 1982-2022 submitted for the 2024 report with the 1982-2023 to be submitted for 2025 because a change to the VAST code caused my previous setup to break.

Also, its a full new input NEFSC food habits dataset, rather than tagging on years as was done in the past. NEAMAP just added one year from last year's submission.

I can no longer produce the inshore <3 nmi and offshore strata which used a custom extrapolation grid. There may be a fix to estimate the bluefish assessment indices ([see here](https://github.com/James-Thorson-NOAA/FishStatsUtils/issues/96)), but the current run uses the built in grid and only estimates EPU level indices for the SOE.

# Workflow

Operational updates to the forage index submitted to the State of the Ecosystem report are in the forageindex github repository: https://github.com/NOAA-EDAB/forageindex

Data input files are in the folder `fhdat` and were processed with the script `VASTforage_ProcessInputDat.R` in that folder: https://github.com/NOAA-EDAB/forageindex/blob/main/fhdat/VASTforage_ProcessInputDat.R

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/forageindex/main/fhdat/VASTforage_ProcessInputDat.R"), eval=FALSE}

```


For the 2024 report, VAST models were run using the script `VASTunivariate_seasonalforageindex_operational.R` in the folder `VASTscripts`: https://github.com/NOAA-EDAB/forageindex/blob/main/VASTscripts/VASTunivariate_seasonalforageindex_operational.R 

That previous script fails for me under VAST v. ‘3.11.2’ and FishStatsUtils v. ‘2.13.1’. I also had to upgrade R to 4.4.1, all of this to run WHAM. 

So, for the 2025 report I have used this script to get our indices in on time: https://github.com/NOAA-EDAB/forageindex/blob/main/VASTscripts/VASTunivariate_seasonalforageindex_SOE_EPUsonly.R 

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/forageindex/refs/heads/main/VASTscripts/VASTunivariate_seasonalforageindex_SOE_EPUsonly.R"), eval=FALSE}

```

Model output was saved in the folder `SOEpyindex` for the time period submitted in 2024 plus the updated data: 1982-2023.

The script to create the SOE forage indices from the VAST output is in the `SOEpyindex` folder: https://github.com/NOAA-EDAB/forageindex/blob/main/SOEpyindex/SOE-VASTForageIndices.R

```{r, code = readLines("https://raw.githubusercontent.com/NOAA-EDAB/forageindex/main/SOEpyindex/SOE-VASTForageIndices.R"), eval=FALSE}

```

Code blocks below setup the index comparisons in the remainder of this document, where I compare only the 1982-2022 (2024 SOE) and 1982-2023 (2025 SOE) indices.

```{r stratasets}
# compare full custom strata set to to EPU only strata

stratlook <- data.frame(Stratum = c("ALLsplit_Stratum_1",
                                    "ALLsplit_Stratum_2",
                                    "ALLsplit_Stratum_3",
                                    "ALLsplit_Stratum_4",
                                    "ALLsplit_Stratum_5",
                                    "ALLsplit_Stratum_6",
                                    "ALLsplit_Stratum_7",
                                    "ALLsplit_Stratum_8",
                                    "ALLsplit_Stratum_9",
                                    "ALLsplit_Stratum_10",
                                    "ALLsplit_Stratum_11",
                                    "ALLsplit_Stratum_12",
                                    "ALLsplit_Stratum_13",
                                    "ALLsplit_Stratum_14",
                                    "ALLsplit_Stratum_15"),
                        Region  = c("AllEPU", 
                                    "MABGB", 
                                    "MABGBstate", 
                                    "MABGBfed", 
                                    "MAB",
                                    "GB",
                                    "GOM",
                                    "bfall",
                                    "bfin",
                                    "bfoff",
                                    "MABGBalbinshore",
                                    "MABGBothoffshore",
                                    "albbfin",
                                    "albbfall",
                                    "allother"))
# new run has fewer strata
stratlook2 <- data.frame(Stratum = c("EPUonly_Stratum_1",
                                     "EPUonly_Stratum_2",
                                     "EPUonly_Stratum_3",
                                     "EPUonly_Stratum_4",
                                     "EPUonly_Stratum_5",
                                     "EPUonly_Stratum_6"),
                         Region = c("AllEPU", 
                                    "MAB",
                                    "GB",
                                    "GOM",
                                    "SS",
                                    "allother"))

stratlookall <- dplyr::bind_rows(stratlook, stratlook2) 

```

```{r getmodinfoindex}

# from each output folder in pyindex, 
outdir <- here("SOEpyindex")
moddirs <- list.dirs(outdir)
# only get pred sensitivities and original full models
fullfall <- grep("/allagg_fall_", moddirs, value = TRUE)
fullspring <- grep("/allagg_spring_", moddirs, value = TRUE)
moddirs <- c(fullfall, fullspring)
# get original published version 1985-2021
#outdir1 <- here("pyindex")
#moddirs1 <- list.dirs(outdir1)
#RTfall <- grep("pyindex/2022/allagg_fall_", moddirs1, value = TRUE)
#RTspring <- grep("pyindex/2022/allagg_spring_", moddirs1, value = TRUE)
#paperfall <- grep("pyindex/allagg_fall_", moddirs1, value = TRUE)
#paperspring <- grep("pyindex/allagg_spring_", moddirs1, value = TRUE)
#moddirs1 <- c(RTfall, RTspring, paperfall, paperspring)
# keep folder name
modnames <- list.dirs(outdir, full.names = FALSE)
#modnames1 <- list.dirs(outdir1, full.names = FALSE)

# function to apply extracting info
getmodinfo <- function(d.name){
  # read settings
  modpath <- stringr::str_split(d.name, "/", simplify = TRUE)
  modname <- modpath[length(modpath)]
  modnamepre <- modpath[length(modpath)-1]
  
  settings <- read.table(file.path(d.name, "settings.txt"), comment.char = "",
    fill = TRUE, header = FALSE)
  
  n_x <- as.numeric(as.character(settings[(which(settings[,1]=="$n_x")+1),2]))
  grid_size_km <- as.numeric(as.character(settings[(which(settings[,1]=="$grid_size_km")+1),2]))
  max_cells <- as.numeric(as.character(settings[(which(settings[,1]=="$max_cells")+1),2]))
  use_anisotropy <- as.character(settings[(which(settings[,1]=="$use_anisotropy")+1),2])
  fine_scale <- as.character(settings[(which(settings[,1]=="$fine_scale")+1),2])
  bias.correct <- as.character(settings[(which(settings[,1]=="$bias.correct")+1),2])
  
  #FieldConfig
  if(settings[(which(settings[,1]=="$FieldConfig")+1),1]=="Component_1"){
    omega1 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+2),2])
    omega2 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+3),1])
    epsilon1 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+4),2])
    epsilon2 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+5),1])
    beta1 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+6),2])
    beta2 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+7),1])
  }
  
  if(settings[(which(settings[,1]=="$FieldConfig")+1),1]=="Omega1"){
    omega1 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+3),1])
    omega2 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+4),1])
    epsilon1 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+3),2])
    epsilon2 <- as.character(settings[(which(settings[,1]=="$FieldConfig")+4),2])
    beta1 <- "IID"
    beta2 <- "IID"
  }
  
  
  #RhoConfig
  rho_beta1 <- as.numeric(as.character(settings[(which(settings[,1]=="$RhoConfig")+3),1]))
  rho_beta2 <- as.numeric(as.character(settings[(which(settings[,1]=="$RhoConfig")+3),2]))
  rho_epsilon1 <- as.numeric(as.character(settings[(which(settings[,1]=="$RhoConfig")+4),1]))
  rho_epsilon2 <- as.numeric(as.character(settings[(which(settings[,1]=="$RhoConfig")+4),2]))
  
  # read parameter estimates, object is called parameter_Estimates
  load(file.path(d.name, "parameter_estimates.RData"))
  
  AIC <- parameter_estimates$AIC[1]  
  converged <- parameter_estimates$Convergence_check[1]
  fixedcoeff <- unname(parameter_estimates$number_of_coefficients[2])
  randomcoeff <- unname(parameter_estimates$number_of_coefficients[3])
  
  #index <- read.csv(file.path(d.name, "Index.csv"))
  
  # return model attributes as a dataframe
  out <- data.frame(modname = paste0(modnamepre,"_", modname),
                    n_x = n_x,
                    grid_size_km = grid_size_km,
                    max_cells = max_cells,
                    use_anisotropy = use_anisotropy,
                    fine_scale =  fine_scale,
                    bias.correct = bias.correct,
                    omega1 = omega1,
                    omega2 = omega2,
                    epsilon1 = epsilon1,
                    epsilon2 = epsilon2,
                    beta1 = beta1,
                    beta2 = beta2,
                    rho_epsilon1 = rho_epsilon1,
                    rho_epsilon2 = rho_epsilon2,
                    rho_beta1 = rho_beta1,
                    rho_beta2 = rho_beta2,
                    AIC = AIC,
                    converged = converged,
                    fixedcoeff = fixedcoeff,
                    randomcoeff = randomcoeff#,
                    #index = index
  )
  	
	return(out)

}

# getfitdat <- function(d.name){
#   # read settings
#   modpath <- stringr::str_split(d.name, "/", simplify = TRUE)
#   modname <- modpath[length(modpath)]
#   
#   readRDS(file.path(d.name, "fit.rds"))
#   nstations <- fit$data_list$n_i # this is grid size not input N stations
#   out <- data.frame(modname = modname,
#                     nstations = nstations
#   )
#   
#   return(out)
# }


# function to apply extracting info
getmodindex <- function(d.name){
  # read settings
  modpath <- stringr::str_split(d.name, "/", simplify = TRUE)
  modname <- modpath[length(modpath)]
  timeseries <- modpath[length(modpath)-1]
  strata <- stringr::str_split(modname, "_", simplify = TRUE)[,5]
  
  if(file.exists(file.path(d.name,"Index.csv"))){
    index <- read.csv(file.path(d.name, "Index.csv"))
    
    # return model indices as a dataframe
    out <- data.frame(modname = paste0(timeseries,"_", modname),
                      strata = strata,
                      index
    )
  }else{
    stopifnot()
    out <- NULL
  }
  
  
  return(out)
}

moddirsall <- c(moddirs)#, moddirs1)

modcompare <- purrr::map_dfr(moddirsall, getmodinfo)

modcompareindex <- purrr::map_dfr(moddirsall, getmodindex)

```

# Check for convergence, etc

```{r}
modtable <- function(moddirs){
  
  # apply getmodinfo function to inout directories
  modcompare <- purrr::map_dfr(moddirs, getmodinfo)
  
  modselect <- modcompare |>
    dplyr::mutate(season = dplyr::case_when(stringr::str_detect(modname, "_fall_") ~ "Fall",
                                            stringr::str_detect(modname, "spring") ~ "Spring",
                                            stringr::str_detect(modname, "_all_") ~ "Annual",
                                            TRUE ~ as.character(NA))) |>
    dplyr::mutate(converged2 = dplyr::case_when(stringr::str_detect(converged, "no evidence") ~ "likely",
                                                stringr::str_detect(converged, "is likely not") ~ "unlikely",
                                                TRUE ~ as.character(NA))) |>
    dplyr::mutate(timeseries = stringr::str_extract(modname, "[^_]+")) |>
    #dplyr::mutate(modname = str_extract(modname, '(?<=allagg_).*')) |>
    dplyr::group_by(timeseries, season) |>
    #dplyr::mutate(deltaAIC = AIC-min(AIC)) |>
    dplyr::select(timeseries, modname, season, #deltaAIC, 
                  fixedcoeff,
                  randomcoeff, use_anisotropy, 
                  omega1, omega2, epsilon1, epsilon2, 
                  beta1, beta2, AIC, converged2) |>
    dplyr::arrange(timeseries, season, AIC)
  
  return(modselect)
}

modselect <- modtable(moddirs)

flextable::flextable(modselect %>%
                       dplyr::mutate(modname = str_extract(modname, '(?<=allagg_).*')) |>
                       dplyr::filter(timeseries %in% c("1982-2022", "1982-2023")) |>
                       dplyr::select(-c(use_anisotropy, 
         omega1, omega2, epsilon1, epsilon2,
         beta1, beta2))) %>%
  flextable::set_header_labels(timeseries = "Data",
                               modname = "Model name",
                               season = "Season",
                               #deltaAIC = "dAIC",
                               fixedcoeff = "N fixed",
                               randomcoeff = "N random",
                               converged2 = "Convergence") |>
  flextable::set_caption("Convergence check for 2025 SOE submission relative to 2024.") %>%
  flextable::fontsize(size = 9, part = "all") %>%
  flextable::colformat_double(digits = 2) |>
  flextable::set_table_properties(layout = "autofit", width = 1)

# # lets only look at indices for converged models
# modcompare_conv <- modselect |>
#   dplyr::ungroup() |>
#   dplyr::filter(converged2 == "likely") |>
#   dplyr::select(modname) |>
#   as.vector() |>
#   unname() |>
#   unlist()
# 
# moddirs_conv <-  moddirs[grepl(sprintf("\\.*(%s)$", paste(modcompare_conv, collapse = '|')), moddirs)]

```


# Trend comparisons

## Methods change and adding one year

Lets compare last year's 2024 SOE submission with the 2025 submission using the same model settings but without the custom extrapolation grid, and with one more year of data (2023).

This is fall trend.

```{r comptrend1, fig.cap="Trend comparison between fall forage indices using 1982-2022 (SOE 2024), and 1982-2023 (SOE 2025) with standard VAST grid.", fig.height=8}
splitoutput <- modcompareindex %>%
  dplyr::mutate(Season = modname |> map(str_split, pattern = "_") |> map_chr(c(1,3))) %>%
  dplyr::mutate(Data = modname |> map(str_split, pattern = "_") |> map_chr(c(1,1))) %>% 
  dplyr::mutate(modname = str_extract(modname, '(?<=allagg_).*')) |>
  tidyr::unite("Stratum", c(strata, Stratum) ) |>
  dplyr::left_join(stratlookall) %>%
  dplyr::filter(Region %in% c("GOM", "GB", "MAB", "AllEPU", "allother")) %>%
  dplyr::mutate(Type = ifelse(Region %in% c("GOM", "GB", "MAB"), "Ecoregion", "Coastwide"))

foragemax <- max(splitoutput$Estimate)


foragetsmean <- splitoutput %>%
  dplyr::group_by(modname, Region) %>%
  dplyr::mutate(fmean = mean(Estimate)) 


ggplot(foragetsmean |> filter(Season =="fall",
                             Data %in% c("1982-2022",
                                         "1982-2023")), 
       aes(x=Time, y=((Estimate-fmean)/fmean), colour=Data)) +
  #geom_errorbar(aes(ymin=(Estimate+Std..Error.for.Estimate - fmean)/fmean, 
  #                  ymax=(Estimate-Std..Error.for.Estimate - fmean)/fmean))+
  geom_point(aes(shape=Data))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  #scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to time series mean")  +
  ggtitle("Fall models: trend comparison") +
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0)
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))

```

This is spring trend.

```{r comptrend2, fig.cap="Trend comparison between spring forage indices using 1982-2022 (SOE 2024), and 1982-2023 (SOE 2025) with standard VAST grid.", fig.height=8}

ggplot(foragetsmean |> filter(Season =="spring",
                             Data %in% c("1982-2022",
                                         "1982-2023")), 
       aes(x=Time, y=((Estimate-fmean)/fmean), colour=Data)) +
  #geom_errorbar(aes(ymin=(Estimate+Std..Error.for.Estimate - fmean)/fmean,
  #                  ymax=(Estimate-Std..Error.for.Estimate - fmean)/fmean))+
  geom_point(aes(shape=Data))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  #scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to time series mean")  +
  ggtitle("Spring models: trend comparison")+
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0)
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))


```

Trends look comparable. The difference looks like the time series mean changing with the addition of 2023. So far so good. 

## Stomach spatial coverage

Spatial coverage for stomachs in fall

```{r}
newmods <- moddirs[stringr::str_detect(moddirs,"EPUonly")]

mapfall <- paste0(newmods[stringr::str_detect(newmods,"_fall_")], "/Data_by_year.png")

knitr::include_graphics(mapfall) 
```

Spatial coverage for stomachs in spring

```{r}
mapspring <- paste0(newmods[stringr::str_detect(newmods,"_spring_")], "/Data_by_year.png")

knitr::include_graphics(mapspring) 
```

# Scale comparisons

We don't use scale much, but in case someone does here they are, scaled relative to the max observed across all datasets.

## Methods change and adding one year

Same comparisons as above: fall

```{r compscale1, fig.cap="Scale comparison between fall forage indices using 1982-2022 (SOE 2024), and 1982-2023 (SOE 2025) with standard VAST grid.", fig.height=8}


ggplot(splitoutput |> filter(Season =="fall",
                             Data %in% c("1982-2022",
                                         "1982-2023")), 
       aes(x=Time, y=Estimate, colour=Data)) +
  #geom_errorbar(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate))+
  geom_ribbon(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate, fill=Data), linetype = 0, alpha = 0.15)+
  geom_point(aes(shape=Data))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to maximum")  +
  ggtitle("Fall models: scale comparison")+
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0)
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))

```

Spring

```{r compscale2, fig.cap="Scale comparison between spring forage indices using 1982-2022 (SOE 2024), and 1982-2023 (SOE 2025) with standard VAST grid.", fig.height=8}
  

ggplot(splitoutput |> filter(Season =="spring",
                             Data %in% c("1982-2022",
                                         "1982-2023")), 
       aes(x=Time, y=Estimate, colour=Data)) +
  #geom_errorbar(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate))+
  geom_ribbon(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate, fill=Data), linetype = 0, alpha = 0.15)+
  geom_point(aes(shape=Data))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Relative forage biomass scaled to maximum")  +
  ggtitle("Spring models: scale comparison")+
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0)
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))


```


The spring comparison is what I was hoping for. I'm unclear on why the fall model is different in scale. 

Plot them all on the reported scale

```{r, fig.height=8}
ggplot(splitoutput |> filter(#Season =="spring",
                             Data %in% c("1982-2022",
                                         "1982-2023")), 
       aes(x=Time, y=Estimate, colour=Data)) +
  #geom_errorbar(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate))+
  geom_ribbon(aes(ymin=Estimate+Std..Error.for.Estimate, ymax=Estimate-Std..Error.for.Estimate, fill=Data), linetype = 0, alpha = 0.15)+
  geom_point(aes(shape=Data))+
  geom_line()+
  #acet_wrap(~fct_relevel(Type, "Ecoregion", "Bluefish"), scales = "free_y") +
  facet_wrap(Region~Season, scales = "free_y", ncol = 2,
              labeller = label_wrap_gen(multi_line=FALSE)) +
  #scale_y_continuous(labels=function(x)round(x/foragemax, digits = 2))+
  ylab("Forage biomass output by VAST")  +
  ggtitle("Output comparison")+
  theme(#legend.position = c(1, 0),
        #legend.justification = c(1, 0)
        legend.position = "bottom",
        legend.text = element_text(size=rel(0.5)),
        legend.key.size = unit(0.5, 'lines'),
        legend.title = element_text(size=rel(0.5)))
```


# Center of gravity exploration

Has forage shifted along the coast?

```{r, eval=FALSE}

library(FishStatsUtils)

getcogVAST <- function(d.name){
  
  fit <- VAST::reload_model(readRDS(paste0(d.name,"/fit.rds")))
  
  dir.create(paste0(d.name,"/test"))
  
  cogout <- FishStatsUtils::plot_range_index(Sdreport = fit$parameter_estimates$SD, 
                                             Report = fit$Report, 
                                             TmbData = fit$data_list,
                                             year_labels = as.numeric(fit$year_labels),
                                             years_to_plot = fit$years_to_plot,
                                             Znames = colnames(fit$data_list$Z_xm),
                                             PlotDir = paste0(d.name,"/test")) #already have plots, will delete this directory
  
  saveRDS(cogout, paste0(d.name,"/cogout.rds"))
  
  unlink(paste0(d.name,"/test"), recursive=TRUE) #removes directory with unneeded plots
  
}

purrr::map(newmods, getcogVAST)

```

Fall center of gravity, forage index, significantly more North and East over time (same as last year)
```{r}

cogoutfall <- readRDS(here::here("SOEpyindex/1982-2023/allagg_fall_500_lennosst_EPUonly_biascorrect/cogout.rds"))

cogdat <- as.data.frame(cogoutfall$COG_Table) |>
  dplyr::mutate(direction = ifelse(m==1, "Eastward", "Northward"))

ggplot2::ggplot(cogdat, ggplot2::aes(x = Year, y = COG_hat)) + 
  ggplot2::geom_point() + 
  ecodata::geom_gls() + 
  ggplot2::labs(y = "Center of gravity, km") +
  ggplot2::facet_wrap(~direction, scales = "free_y") + 
  ecodata::theme_facet()

```

Spring center of gravity, forage index, significantly northward with the addition of 2023 data (no significant trends last year)
```{r}

cogoutspring <- readRDS(here::here("SOEpyindex/1982-2023/allagg_spring_500_lennosst_EPUonly_biascorrect/cogout.rds"))

cogdat <- as.data.frame(cogoutspring$COG_Table) |>
  dplyr::mutate(direction = ifelse(m==1, "Eastward", "Northward"))

ggplot2::ggplot(cogdat, ggplot2::aes(x = Year, y = COG_hat)) + 
  ggplot2::geom_point() + 
  ecodata::geom_gls() + 
  ggplot2::labs(y = "Center of gravity, km") +
  ggplot2::facet_wrap(~direction, scales = "free_y") + 
  ecodata::theme_facet()

```

update get function for ecodata to add cog to existing forage_index dataset
```{r, eval=FALSE}
## Forage Index
raw.dir<- here::here("data-raw/")
#ann<- "annualforageindex - Sarah Gaichas - NOAA Federal.rds"
fal<- "fallforageindex - Sarah Gaichas - NOAA Federal.rds"
spr<- "springforageindex - Sarah Gaichas - NOAA Federal.rds"
falcog <- "fallforagecog.rds"
sprcog <- "springforagecog.rds"

get_forage_index <- function(save_clean = F){

  #annual<-readRDS(file.path(raw.dir, ann))
  fall<- readRDS(file.path(raw.dir, fal))
  spring<-readRDS(file.path(raw.dir, spr))
  fallcog <- readRDS(file.path(raw.dir, falcog))
  springcog <-readRDS(file.path(raw.dir, sprcog))

  forage_index<- rbind(fall, spring, fallcog, springcog)


  if (save_clean){
    usethis::use_data(forage_index, overwrite = T)
  } else {
    return(forage_index)
  }
}
get_forage_index(save_clean = T)


```

update plot function
```{r, eval=FALSE}

plot_forage_index <- function(shadedRegion = NULL,
                              report="MidAtlantic",
                              varName = "index") {

  setup <- ecodata::plot_setup(shadedRegion = shadedRegion,
                               report=report)
  
  if (varName == "index") {
    
  if (report == "MidAtlantic") {
    filterEPUs <- c("MAB")
  } else {
    filterEPUs <- c("GB", "GOM")
  }

  fix<- ecodata::forage_index |>
    dplyr::filter(Var %in% c("Fall Forage Fish Biomass Estimate",
                             "Spring Forage Fish Biomass Estimate"),
                  EPU %in% filterEPUs) |>
    dplyr::group_by(EPU) |>
    dplyr::summarise(max = max(Value))

  p <- ecodata::forage_index |>
    dplyr::filter(Var %in% c("Fall Forage Fish Biomass Estimate",
                             "Fall Forage Fish Biomass Estimate SE",
                             "Spring Forage Fish Biomass Estimate",
                             "Spring Forage Fish Biomass Estimate SE"),
                  EPU %in% filterEPUs) |>
    dplyr::group_by(EPU) |>
    tidyr::separate(Var, into = c("Season", "A", "B", "C", "D", "Var")) |>
    dplyr::mutate(Var = tidyr::replace_na(Var, "Mean")) |> #,
                  #max = as.numeric(Value)) |>
    tidyr::pivot_wider(names_from = Var, values_from = Value) |>
    dplyr::left_join(fix) |>
    dplyr::mutate(#Value = Value/resca,
      Mean = as.numeric(Mean),
      #max = as.numeric(Value),
      Mean = Mean/max,
      SE = SE/max,
      Upper = Mean + SE,
      Lower = Mean - SE) |>
    ggplot2::ggplot(ggplot2::aes(x = Time, y = Mean, group = Season))+
    ggplot2::annotate("rect", fill = setup$shade.fill, alpha = setup$shade.alpha,
        xmin = setup$x.shade.min , xmax = setup$x.shade.max,
        ymin = -Inf, ymax = Inf) +
    ggplot2::geom_ribbon(ggplot2::aes(ymin = Lower, ymax = Upper, fill = Season), alpha = 0.5)+
    ggplot2::geom_point()+
    ggplot2::geom_line()+
    ggplot2::ggtitle("")+
    ggplot2::ylab(expression("Relative forage biomass"))+
    ggplot2::xlab(ggplot2::element_blank())+
    ggplot2::facet_wrap(.~EPU)+
    ecodata::geom_gls()+
    ecodata::theme_ts()+
    ecodata::theme_facet()+
    ecodata::theme_title()

    if (report == "NewEngland") {
      p <- p +
        ggplot2::theme(legend.position = "bottom",
                       legend.title = ggplot2::element_blank())

    }
  }
  
  if (varName == "cog"){
    
    p <- ecodata::forage_index |>
      dplyr::filter(Var %in% c("Fall Eastward Forage Fish Center of Gravity",
                               "Fall Eastward Forage Fish Center of Gravity SE",
                               "Fall Northward Forage Fish Center of Gravity",
                               "Fall Northward Forage Fish Center of Gravity SE",
                               "Spring Eastward Forage Fish Center of Gravity",
                               "Spring Eastward Forage Fish Center of Gravity SE",
                               "Spring Northward Forage Fish Center of Gravity",
                               "Spring Northward Forage Fish Center of Gravity SE")) |>
      tidyr::separate(Var, into = c("Season", "Direction", "A", "B", "C", "D", "E", "Var")) |>
      dplyr::mutate(Var = tidyr::replace_na(Var, "Mean")) |> 
      tidyr::pivot_wider(names_from = Var, values_from = Value) |>
      dplyr::mutate(Mean = as.numeric(Mean),
                    Upper = Mean + SE,
                    Lower = Mean - SE) |>
      ggplot2::ggplot(ggplot2::aes(x = Time, y = Mean, group = Season))+
      ggplot2::annotate("rect", fill = setup$shade.fill, alpha = setup$shade.alpha,
                        xmin = setup$x.shade.min , xmax = setup$x.shade.max,
                        ymin = -Inf, ymax = Inf) +
      ggplot2::geom_ribbon(ggplot2::aes(ymin = Lower, ymax = Upper, fill = Season), alpha = 0.3)+ #
      ggplot2::geom_point()+
      ggplot2::geom_line()+
      ggplot2::ggtitle("")+
      ggplot2::ylab(expression("Forage Center of Gravity, km"))+
      ggplot2::xlab(ggplot2::element_blank())+
      ggplot2::facet_wrap(~Direction, scales = "free_y")+ #Season
      ecodata::geom_gls()+
      ecodata::theme_ts()+
      ecodata::theme_facet()+
      ecodata::theme_title()

  }

    return(p)

}

attr(plot_forage_index,"report") <- c("MidAtlantic","NewEngland")
attr(plot_forage_index, "varName") <- c("index", "cog")

```

forage COG indices produced by SOE-VASTForageCOG.R

```{r}

fallforagecog <- readRDS(here::here("SOEpyindex/fallforagecog.rds"))
springforagecog <- readRDS(here::here("SOEpyindex/springforagecog.rds"))

setup <- ecodata::plot_setup(shadedRegion = c(2014,2023), report = "MidAtlantic")

testplot <- dplyr::bind_rows(fallforagecog, springforagecog) |>
  
  dplyr::filter(Var %in% c("Fall Eastward Forage Fish Center of Gravity",
                               "Fall Eastward Forage Fish Center of Gravity SE",
                               "Fall Northward Forage Fish Center of Gravity",
                               "Fall Northward Forage Fish Center of Gravity SE",
                               "Spring Eastward Forage Fish Center of Gravity",
                               "Spring Eastward Forage Fish Center of Gravity SE",
                               "Spring Northward Forage Fish Center of Gravity",
                               "Spring Northward Forage Fish Center of Gravity SE")) |>
      tidyr::separate(Var, into = c("Season", "Direction", "A", "B", "C", "D", "E", "Var")) |>
      dplyr::mutate(Var = tidyr::replace_na(Var, "Mean")) |> 
      tidyr::pivot_wider(names_from = Var, values_from = Value) |>
      dplyr::mutate(Mean = as.numeric(Mean),
                    Upper = Mean + SE,
                    Lower = Mean - SE) |>
      ggplot2::ggplot(ggplot2::aes(x = Time, y = Mean, group = Season))+
      ggplot2::annotate("rect", fill = setup$shade.fill, alpha = setup$shade.alpha,
                        xmin = setup$x.shade.min , xmax = setup$x.shade.max,
                        ymin = -Inf, ymax = Inf) +
      ggplot2::geom_ribbon(ggplot2::aes(ymin = Lower, ymax = Upper, fill = Season), alpha = 0.3)+ #
      ggplot2::geom_point()+
      ggplot2::geom_line()+
      ggplot2::ggtitle("")+
      ggplot2::ylab(expression("Forage Center of Gravity, km"))+
      ggplot2::xlab(ggplot2::element_blank())+
      ggplot2::facet_wrap(~Direction, scales = "free_y")+ #Season
      ecodata::geom_gls()+
      ecodata::theme_ts()+
      ecodata::theme_facet()+
      ecodata::theme_title()

testplot

```

